### 2025-02-23 rebuttal.tex

\documentclass[10pt,twocolumn,letterpaper]{article}
\usepackage[rebuttal]{cvpr}

% Include other packages here, before hyperref.
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}

% Import additional packages in the preamble file, before hyperref
\input{preamble}

% If you comment hyperref and then uncomment it, you should delete
% egpaper.aux before re-running latex.  (Or just hit 'q' on the first latex
% run, let it finish, and you should be clear).
\definecolor{cvprblue}{rgb}{0.21,0.49,0.74}
\usepackage[pagebackref,breaklinks,colorlinks,allcolors=cvprblue]{hyperref}

% If you wish to avoid re-using figure, table, and equation numbers from
% the main paper, please uncomment the following and change the numbers
% appropriately.
%\setcounter{figure}{2}
%\setcounter{table}{1}
%\setcounter{equation}{2}

% If you wish to avoid re-using reference numbers from the main paper,
% please uncomment the following and change the counter value to the
% number of references you have in the main paper (here, 100).
%\makeatletter
%\apptocmd{\thebibliography}{\global\c@NAT@ctr 100\relax}{}{}
%\makeatother

%%%%%%%%% PAPER ID  - PLEASE UPDATE
\def\paperID{*****} % *** Enter the Paper ID here
\def\confName{CVPR}
\def\confYear{2025}

\begin{document}

%%%%%%%%% TITLE - PLEASE UPDATE
\title{\LaTeX\ Leveraging Diffusion Based Image Generation to Investigate its Impact on Single Cell Detection Accuracy}  % **** Enter the paper title here

\maketitle
\thispagestyle{empty}
\appendix

%%%%%%%%% BODY TEXT - ENTER YOUR RESPONSE BELOW
\section{Introduction}

After receiving paper reviews, authors may optionally submit a rebuttal to address the reviewers' comments, which will be limited to a {\bf one page} PDF file.
Please follow the steps and style guidelines outlined below for submitting your author response.

The author rebuttal is optional and, following similar guidelines to previous conferences, is meant to provide you with an opportunity to rebut factual errors or to supply additional information requested by the reviewers.
It is NOT intended to add new contributions (theorems, algorithms, experiments) that were absent in the original submission and NOT specifically requested by the reviewers.
You may optionally add a figure, graph, or proof to your rebuttal to better illustrate your answer to the reviewers' comments.

Per a passed 2018 PAMI-TC motion, reviewers should refrain from requesting significant additional experiments for the rebuttal or penalize for lack of additional experiments.
Authors should refrain from including new experimental results in the rebuttal, especially when not specifically requested to do so by the reviewers.
Authors may include figures with illustrations or comparison tables of results reported in the submission/supplemental material or in other papers.

Just like the original submission, the rebuttal must maintain anonymity and cannot include external links that reveal the author identity or circumvent the length restriction.
The rebuttal must comply with this template (the use of sections is not required, though it is recommended to structure the rebuttal for ease of reading).

%-------------------------------------------------------------------------

\subsection{Response length}
Author responses must be no longer than 1 page in length including any references and figures.
Overlength responses will simply not be reviewed.
This includes responses where the margins and formatting are deemed to have been significantly altered from those laid down by this style guide.
Note that this \LaTeX\ guide already sets figure captions and references in a smaller font.

%------------------------------------------------------------------------
\section{Formatting your Response}

{\bf Make sure to update the paper title and paper ID in the appropriate place in the tex file.}

All text must be in a two-column format.
The total allowable size of the text area is $6\frac78$ inches (17.46 cm) wide by $8\frac78$ inches (22.54 cm) high.
Columns are to be $3\frac14$ inches (8.25 cm) wide, with a $\frac{5}{16}$ inch (0.8 cm) space between them.
The top margin should begin 1 inch (2.54 cm) from the top edge of the page.
The bottom margin should be $1\frac{1}{8}$ inches (2.86 cm) from the bottom edge of the page for $8.5 \times 11$-inch paper;
for A4 paper, approximately $1\frac{5}{8}$ inches (4.13 cm) from the bottom edge of the page.

Please number any displayed equations.
It is important for readers to be able to refer to any particular equation.

Wherever Times is specified, Times Roman may also be used.
Main text should be in 10-point Times, single-spaced.
Section headings should be in 10 or 12 point Times.
All paragraphs should be indented 1 pica (approx.~$\frac{1}{6}$ inch or 0.422 cm).
Figure and table captions should be 9-point Roman type as in \cref{fig:onecol}.

List and number all bibliographical references in 9-point Times, single-spaced,
at the end of your response.
When referenced in the text, enclose the citation number in square brackets, for example~\cite{Alpher05}.
Where appropriate, include the name(s) of editors of referenced books.

\begin{figure}[t]
  \centering
  \fbox{\rule{0pt}{0.5in} \rule{0.9\linewidth}{0pt}}
  %\includegraphics[width=0.8\linewidth]{egfigure.eps}
   \caption{Example of caption.  It is set in Roman so that mathematics
   (always set in Roman: $B \sin A = A \sin B$) may be included without an
   ugly clash.}
   \label{fig:onecol}
\end{figure}

To avoid ambiguities, it is best if the numbering for equations, figures, tables, and references in the author response does not overlap with that in the main paper (the reviewer may wonder if you talk about \cref{fig:onecol} in the author response or in the paper).
See \LaTeX\ template for a workaround.

%-------------------------------------------------------------------------
\subsection{Illustrations, graphs, and photographs}

All graphics should be centered.
Please ensure that any point you wish to make is resolvable in a printed copy of the response.
Resize fonts in figures to match the font in the body text, and choose line widths which render effectively in print.
Readers (and reviewers), even of an electronic copy, may choose to print your response in order to read it.
You cannot insist that they do otherwise, and therefore must not assume that they can zoom in to see tiny details on a graphic.

When placing figures in \LaTeX, it is almost always best to use \verb+\includegraphics+, and to specify the  figure width as a multiple of the line width as in the example below
{\small\begin{verbatim}
   \usepackage{graphicx} ...
   \includegraphics[width=0.8\linewidth]
                   {myfile.pdf}
\end{verbatim}
}

%%%%%%%%% REFERENCES
{
    \small
    \bibliographystyle{ieeenat_fullname}
    \bibliography{main}
}

\end{document}

### 2025-02-26 sec/1_introduction.tex
extend introduction, related works and methodology chapters

\section{Introduction}
\label{sec:introduction}
Single cell detection in microscopy images is a fundamental task in biological and medical applications, enabling quantitative analysis of cellular processes and disease mechanisms~\cite{meijering_cell_2012}.
Brightfield microscopy, a label-free technique, is widely used but presents challenges for automated analysis due to low contrast and variability in cell appearance~\cite{jyrki_selinummi_bright_2009}.
Deep learning, particularly CNN-based object detectors, has shown great promise for automatic cell detection~\cite{erick_moen_deep_2019,thorsten_falk_u-net_2019}.
However, the performance of these models is heavily dependent on large, annotated datasets, which are expensive and time-consuming to acquire in microscopy~\cite{ronneberger_u-net_2015}.

Synthetic data generation offers a potential solution to alleviate data scarcity.
Recent advancements in diffusion models have demonstrated their ability to generate high-fidelity images across various domains~\cite{ho_denoising_2020,song_denoising_2020}.
In microscopy, synthetic data could augment real datasets, improve model generalization, and reduce annotation efforts~\cite{rajaram_simucell_2012,trampert_deep_2021,lehmussola_synthetic_2008}.

This paper explores the use of unconditional diffusion models for generating synthetic brightfield microscopy images and investigates their impact on single cell detection accuracy.
We address the central research question: \textbf{Can synthetic brightfield microscopy images generated by diffusion models enhance the performance of object detection models for single cell detection?}
We further investigate: (1) the perceptual realism of generated images through expert evaluation, and (2) the influence of synthetic data proportion in training datasets.

>>>>

\section{Introduction}
\label{sec:introduction}
Single cell detection in microscopy images is a fundamental task in biological and medical applications, enabling quantitative analysis of cellular processes and disease mechanisms~\cite{meijering_cell_2012}.
Accurate identification and localization of individual cells are crucial for understanding cellular heterogeneity, dynamics, and responses to stimuli, driving advancements in fields ranging from drug discovery to diagnostics~\cite{f_mualla_automatic_2016,e_d_ferreira_classification_2024}.
Brightfield microscopy, a label-free technique, is widely used due to its non-invasiveness and simplicity, allowing for long-term observation of live cells without phototoxicity or alteration of cellular states~\cite{jyrki_selinummi_bright_2009,huixia_ren_cellbow_2020}.
However, brightfield microscopy presents significant challenges for automated analysis due to inherent limitations such as low contract between cells and background, high variability in cell appearance influenced by factors like cell type and imaging conditions, and the presence of artifacts like uneven illumination and well edge distortions~\cite{jyrki_selinummi_bright_2009,f_mualla_automatic_2016}.
These challenges often necessitate labor-intensive manual annotation, hindering high-throughput analysis and objective analysis.

Deep learning, particularly CNN-based object detectors, has shown great promise for automatic cell detection~\cite{erick_moen_deep_2019,thorsten_falk_u-net_2019}.
However, the performance of these models is heavily dependent on large, annotated datasets, which are expensive and time-consuming to acquire in microscopy~\cite{ronneberger_u-net_2015}.
Moreover, ethical considerations and standardization issues further complicate the acquisition of sufficient real-world data, especially for \textit{in vitro} studies involving diverse cell lines and experimental conditions.

Synthetic data generation offers a potential solution to alleviate data scarcity.
Recent advancements in diffusion models have demonstrated their ability to generate high-fidelity images across various domains~\cite{ho_denoising_2020,song_denoising_2020}.
In microscopy, synthetic data could augment real datasets, improve model generalization, and reduce annotation efforts~\cite{rajaram_simucell_2012,trampert_deep_2021,lehmussola_synthetic_2008}.
Diffusion models, unlike GANs, offer improved training stability and the ability to generate diverse samples, making them particularly attractive for capturing the inherent variability in cell morphology and imaging artifacts often encountered in brightfield microscopy.

This paper explores the use of unconditional diffusion models for generating synthetic brightfield microscopy images of CHO cells and investigates their impact on single cell detection accuracy using state-of-the-art object detectors.
The diffusion model was trained on a carefully curated dataset of 10,000 images (512x512 pixels) of stably transfected CHO-K1 and CHO DG44 cell lines, acquired using a high-throughput microscope.
We address the central research question: \textbf{Can synthetic brightfield microscopy images generated by diffusion models enhance the performance of object detection models for single cell detection?}
We further investigate: (1) the perceptual realism of generated images through expert evaluation, and (2) the influence of synthetic data proportion in training datasets.

### 2025-02-26 sec/2_relatedwork.tex
extend introduction, related works and methodology chapters

\section{Related Work}
\label{sec:related-work}
Synthetic microscopy image generation has gained traction for data augmentation and algorithm development.
Early methods employed rule-based models and physical simulations~\cite{david_svoboda_generation_2012,david_svoboda_towards_2013,rajaram_simucell_2012}, often limited by their ability to capture complex biological diversity.

Deep learning-based generative models, particularly GANs and diffusion models, have emerged as powerful alternatives.
GANs have been used for cross-modality translation~\cite{christiansen_silico_2018,gyuhyun_lee_deephcs_2018,gyuhyun_lee_deephcs_2021}, super-resolution~\cite{zhang_high-throughput_2019}, and cell image generation~\cite{marin_scalbert_generic_2019}.
However, GAN training can be unstable and prone to mode collapse~\cite{juan_c_caicedo_evaluation_2019,ricard_durall_combating_2020}.

Diffusion models, inspired by non-equilibrium thermodynamics~\cite{jascha_sohl-dickstein_deep_2015}, offer improved training stability and sample quality~\cite{ho_denoising_2020,song_denoising_2020}.
Applications in microscopy include painting image generation~\cite{cross-zamirski_class-guided_2023}, super-resolution~\cite{gabriel_della_maggiora_conditional_2023}, electron microscopy enhancement~\cite{lu_diffusion-based_2024}, and physics-informed reconstruction~\cite{li_microscopy_2023}.

While diffusion models are increasingly used in microscopy image generation, research on their direct application for enhancing \textbf{object detection} in brightfield microscopy remains limited.
This work addresses this gap by systematically evaluating the impact of diffusion-based synthetic brightfield microscopy image data on single cell detection performance using state-of-the-art object detection models.

>>>>

\section{Related Work}
\label{sec:related-work}
Synthetic microscopy image generation has become an increasingly important area of research, driven by the need for large, annotated datasets for training and validating image analysis algorithms.
Generating synthetic data offers a promising avenue to overcome the limitations of acquiring and annotating real microscopy images, which are often expensive, time-consuming, and require expert knowledge.
Early approaches to synthetic microscopy image generation relied on rule-based models and physical simulations~\cite{david_svoboda_generation_2012,david_svoboda_towards_2013,rajaram_simucell_2012}.
These methods, while offering control over image parameters and ground truth generation, were often limited in their ability to capture the full complexity and biological diversity observed in real-world microscopy data.

Early works in synthetic microscopy image generation focused on creating controlled datasets for algorithm development and validation.
Svoboda \etal pioneered in the generation of fully 3D synthetic brightfield microscopy time-lapse sequences, modeling cell shapes, structures, and motion~\cite{david_svoboda_generation_2012}.
This work was further extended to simulate realistic cell population distributions in 3D by controlling cell count and clustering probability~\cite{david_svoboda_towards_2013}.
SimuCell, developed by Rajaram \etal~\cite{rajaram_simucell_2012}, is another example of a simulation-based approach, aiming to mimic brightfield microscopy images for cell segmentation algorithm development.
These methods, based on predefined rules and physical models, provided valuable initial datasets for testing algorithms, particularly for segmentation and tracking.
However, their inherent limitations in replicating the intricate textures, subtle variations, and artifacts present in real biological images restricted their utility for training more sophisticated, data-driven models, especially deep learning approaches.

The advent of deep learning, particularly Generative Adversarial Networks (GANs), marked a significant shift in synthetic image generation.
GANs offered the potential to learn complex data distributions directly from real images, leading to more realistic and diverse synthetic ouputs.
In the context of microscopy, GANs have been extensively explored for various tasks.
Cross-modality translation, aiming to predict fluorescence microscopy images from brightfield inputs, has been a prominent application.
Christiansen \etal introduced ``In Silico Labeling'' (ISL) using deep learning to predict fluorescence labels from transmitted light images~\cite{christiansen_silico_2018}.
Building upon this, Lee \etal developed DeepHCS and DeepHCS++ to transform brightfield images into multiple fluorescence channels relevant for high-contrast screening, leveraging adversarial losses for realism~\cite{gyuhyun_lee_deephcs_2018,gyuhyun_lee_deephcs_2021}.
GANs have also been applied to enhance image quality achieving super-resolution under a large field-of-view~\cite{zhang_high-throughput_2019} and generate isolate cell images approximating the shape and texture distributions of cell components~\cite{marin_scalbert_generic_2019}.
Despite these success, GANs are known to suffer from training instability and mode collapse, limiting the diversity and reliability of generated samples, particularly when training data is scarce, complex or imbalanced, as often encountered in microscopy~\cite{juan_c_caicedo_evaluation_2019,ricard_durall_combating_2020}.

Diffusion models, inspired by non-equilibrium thermodynamics and rooted in denoising score matching, offer improved training stability, sample quality and diversity~\cite{jascha_sohl-dickstein_deep_2015,ho_denoising_2020,song_denoising_2020}.
These models have demonstrated state-of-the-art performance in various image generation tasks and are increasingly being explored in microscopy.
Cross-Zamirski \etal introduced a class-guided diffusion model to generate cell painting images from brightfield data, along with class labels, demonstrating the potential for controlled image synthesis~\cite{cross-zamirski_class-guided_2023}.
Della Maggiora \etal presented a conditional variational diffusion model for super-resolution microscopy and quantitative phase imaging, focusing on improving the noise schedule learning during training~\cite{gabriel_della_maggiora_conditional_2023}.
In electron microscopy, Lu \etal developed EMDiffuse, a suite of algorithms based on diffusion models for image enhancement, reconstruction, and generation, showcasing the versatility of these models across different microscopy modalities~\cite{lu_diffusion-based_2024}.
Li \etal proposed a physics-informed diffusion model (PI-DDPM) that incorporates the physical model of microscopy image formation into the loss function, leading to improved reconstructions with reduced artifacts compared to regular DDPMs when trained on synthetic microscopy data~\cite{li_microscopy_2023}.
These works highlight the growing interest in diffusion models for various microscopy image synthesis and enhancement tasks, demonstrating their capability to generate high-quality and diverse samples.

While diffusion models are increasingly used in microscopy image generation, research on their direct application for enhancing \textbf{object detection}, specifically in the challenging domain of brightfield microscopy, remains significantly limited.
Existing studies primarily focus on image generation quality evaluation and downstream tasks like segmentation or image-based profiling.
The critical question of whether and how diffusion-based synthetic brightfield microscopy images can effectively improve the performance of object detection models for single cell detection has not been systematically addressed.
This work directly addresses this gap by undertaking a comprehensive evaluation of the impact on diffusion-based synthetic brightfield microscopy image data on single cell detection performance.
By focusing on object detection, this research contributes a novel perspective to the field, exploring the practical utility of synthetic data for a fundamental task in quantitative microscopy.

### 2025-02-26 sec/3_methodology.tex
extend introduction, related works and methodology chapters

\section{Methodology}
\label{sec:methodology}
Our methodology comprises two main stages: (1) unconditional diffusion model training and synthetic image generation, and (2) evaluation of object detection models trained with varying proportions of synthetic data.

\subsection{Image Generation with Diffusion Models}
\label{subsec:image-generation-with-diffusion-models}
We trained an unconditional diffusion model with a U-Net architecture~\cite{ronneberger_u-net_2015} using a dataset of  10,000 patches ($512 \times 512$) extracted from real brightfield microscopy images of stably transfected CHO cells (see supplementary material for dataset details).
The U-Net architecture consisted of down and up-sampling blocks with residual connections and optionally attention mechanisms in the down and up-sampling paths.
We employed the DDIM scheduler~\cite{song_denoising_2020} for efficient sampling and trained the model for 350 epochs using the AdamW optimizer with a learning rate of $1e^{-4}$.
Model selection was based on visual inspection of generated samples and the Fréchet Inception Distance (FID) scores calculated against a held-out set of real images.
The final generation process used the Euler a scheduler with trailing timestep spacing and epsilon prediction type, generating images with inference steps randomized between 35 and 40.

\subsection{Cell Detection Model Training and Evaluation}
\label{subsec:cell-detection-model-training-and-evaluation}
We created six datasets containing synthetic images and a baseline dataset containing 5,000 real images (\textit{scc\_real}), that is used to compare the performance of the detection models trained with synthetic data.
Three datasets used varying proportions of synthetic images (10\%, 30\%, and 50\%) to replace real images in the training set (\textit{scc\_10}, \textit{scc\_30}, \textit{scc\_50}), while the remaining three datasets used the same proportions to add synthetic images to the training set (\textit{scc\_add\_10}, \textit{scc\_add\_30}, \textit{scc\_add\_50}).
Real images were labeled using a semi-automated approach combining fluorescence channel information and manual verification.
Synthetic images were labeled using model-assisted labeling with a pre-trained and fine-tuned YOLOv8m model and manual label refinement.

We fine-tuned three state-of-the-art object detection models: YOLOv8 (sizes s, m, x)~\cite{jocher_ultralytics_2023}, YOLOv9 (sizes c, e)~\cite{Wang2024YOLOv9LW}, and RT-DETR (sizes l, x)~\cite{Lv2023DETRsBY}, pre-trained on COCO~\cite{tsung-yi_lin_microsoft_2014}.
Models were trained for 200 epochs using default Ultralytics augmentation settings and evaluated on a held-out test set of 16,758 real images.
Performance was measured using mean Average Precision (mAP) at IoU thresholds of 0.5 (mAP\@50), 0.75 (mAP\@75), and averaged across IoUs from 0.5 to 0.95 (mAP\@50:95).

\subsection{Expert Survey for Image Realism}
\label{subsec:expert-survey-for-image-realism}
To assess the perceptual realism of generated images, we conducted a survey with 11 microscopy experts and biologists from Synentec GmbH\@.
Participants were presented with 30 randomly ordered images (20 synthetic, 10 real) and asked to classify each as real or generated and rate their confidence.
We analyzed classification accuracy and collected textual explanations for misclassifications to understand the visual cues experts relied upon.

>>>>

\section{Methodology}
\label{sec:methodology}
Our methodology is structured into three distinct yet interconnected stages.
These stages are: (1) training and unconditional diffusion model and generating synthetic brightfield microscopy images, (2) training and evaluating state-of-the-art object detection models using two sets of datasets with varying proportions of synthetic images replacing and adding real images, and
(3) conducting an expert survey to assess the perceptual realism of generated images.
This multi-faceted approach allows for a comprehensive investigation into the efficacy of synthetic data augmentation in the context of cell detection.

\subsection{Image Generation with Diffusion Models}
\label{subsec:image-generation-with-diffusion-models}
The first stage of our methodology focuses on generating realistic synthetic brightfield microscopy images using an unconditional diffusion model.
This stage involves dataset preparation, diffusion model training, model selection, and the final image generation process.

The diffusion model was trained using a dataset of 10,000 patches ($512 \times 512$) extracted from real brightfield microscopy images.
These images were acquired using a CELLAVISTA 3.1 RS HE high-throughput microscope (Synentec GmbH) with a $10\times$ objective lense, resuting in 8-bit grayscale images.
The source material consisted of stably transfected CHO cells (CHO-K1 and CHO DG44 variants) expressing eGFP, imaged in 96-well plates at three different densities (300, 150, and 1 cell per well).

Prior to training, raw subwell images ($3056 \times 3056$ pixels) were patchified into $512 \times 512$ pixel non-overlapping patches to increase the dataset size and focus the model on local cell features.
Patches containing visible well edges, identified as dark arch artifacts due to light refraction at the well boundaries, were manually excluded to ensure the training data primarily comprised relevant cellular structures.
The final dataset of 10,000 patches was randomly sampled from the filtered set of approximately 24,000 patches to manage computation resources within the scope of this research.
For efficient data loading and processing during training, the dataset was converted into the Hugging Face Datasets format, stored as parquet files, which facilitates faster data access and management~\cite{vohra_apache_2016}.

We employed an unconditional diffusion models with a U-Net architectures and attention mechanisms as the backbone for image generation.
The U-Net architecture was chosen for its proven effectiveness in image generation and segmentation tasks, particularly in biomedical imaging, due to its ability to capture both local and global context through its encode-decoder structure and skip connections~\cite{ronneberger_u-net_2015}.
The specific U-Net architecture implemented consisted of downsampling and upsampling blocks with residual connections to facilitate gradient flow and improve training stability.
Attention mechanisms were optionally incorporated into both the downsampling and upsampling paths in some model variants to investigate their impact on image generation quality.
The number of up and down blocks, channel configurations and the placement of attention blocks were varied across different model configurations to evaluate architectural choices.

The diffusion model was trained using the DDIM scheduler~\cite{song_denoising_2020} for efficient sampling during training and to explore faster inference capabilities.
The model was trained for 350 epochs using the AdamW optimizer with a learning rate of $1e^{-4}$.
Moving Average (EMA) with a decay of 0.9999 was applied to the model weights to stabilize training and potentially improve generalization.
A batch size of 4 was used, constrained by GPU memory limitations.
Training was performed using Distributed Data Parallel (DDP) across two NVIDIA A100-SXM4-40GB GPUs with mixed precision (float16) enabled using PyTorch Lightning~\cite{falcon_pytorchlightningpytorch-lightning_2020} to accelerate training.

Model selection was based on a combination of visual inspection of generated samples and quantitative evaluation using the Fréchet Inception Distance (FID) scores.
FID scores were calculated against a held-out set of real images to objectively measure the similarity between the distributions of real and generated images and computed every 10 epochs to track training progress and identify optimal model states.

For the final synthetic image generation, the Euler a scheduler was chosen for its balance of sample quality and generation speed, determined through empirical testing of various schedulers.
Trailing timestep spacing and epsilon prediction type were used based on hyperparameter optimization experiments that indicated these settings produced visually superior results for our dataset.
To introduce variability and potentially enhance the robustness of the downstream object detectors, the number of inference steps during the final generation process was randomized between 35 and 40 steps for each batch of 16 images.
A total of 10,000 synthetic images were generated using this optimized configuration to create datasets for object detection model training.

\subsection{Cell Detection Model Training and Evaluation}
\label{subsec:cell-detection-model-training-and-evaluation}
The second stage of our methodology involved training and evaluating state-of-the-art object detection models using datasets incorporating varying proportions of synthetic images.
This stage aimed to quantify the impact of synthetic data augmentation on cell detection performance.

We created six datasets containing synthetic images and a baseline dataset consisting of 5,000 real images (\textit{scc\_real}).
The baseline dataset served as a control to evaluate the performance of detection models trained solely on real data.
To investigate the effect of replacing real images with synthetic ones, three datasets (\textit{scc\_10}, \textit{scc\_30}, \textit{scc\_50}) were created where 10\%, 30\%, and 50\% of the real images in the training set where replaced with synthetic images, respectively.
\textit{scc\_add\_10}, \textit{scc\_add\_30}, \textit{scc\_add\_50} were created with the same proportions of synthetic images added to the training set, keeping the original real images intact and increasing the total dataset size.

The remaining images from the real dataset (beyond the 5,000 training images) were allocated to fixed validation (2,527 images) and test (16,758 images) sets, ensuring consistent evaluation across all training configurations.
These large validation and test sets kept constant across all experiments to provide robust and statistically meaningful performance comparisons.

Real images were labeled using a semi-automated approach that combined fluorescence channel information with manual verification.
First, a classical image processing algorithm was implemented to automatically detect cells in the fluorescence channel (eGFP signal).
This algorithm involved binarizing the fluorescence image, finding contours corresponding to fluorescent signals, and generating bounding boxes around these contours.
The automatically generated bounding boxes were then manually reviewed and refined by a human expert using the Roboflow annotation tool~\cite{dwyer_roboflow_nodate} to ensure accuracy and correct any errors from the automated process.
This hybrid approach significantly reduced the manual labeling effort while maintaining high-quality annotations.

Synthetic images were labeled using a model-assisted labeling approach.
A YOLOv8m model was initially pre-trained and then fine-tuned on a subset of real labeled images.
This fine-tuned YOLOv8m model was then integrated into the Roboflow annotation tool's ``Model-Assisted Labeling'' feature.
These model-predicted bounding boxes were then manually reviewed and refined by a human expert to correct any inaccuracies, missed detections or false positives.
This approach leveraged the pre-trained object detection model to expedite the labeling of synthetic data, while manual refinement ensured the final labels were of high quality.

We fine-tuned three state-of-the-art object detection model families: YOLOv8 (sizes s, m, x)~\cite{jocher_ultralytics_2023}, YOLOv9 (sizes c, e)~\cite{Wang2024YOLOv9LW}, and RT-DETR (sizes l, x)~\cite{Lv2023DETRsBY}.
These models were chosen for their state-of-the-art performance, diverse architectures (CNN-based and Transformer-based), and availability within the Ultralytics framework~\cite{jocher_ultralytics_2023}, which facilitated consistent training and evaluation.
All models were pre-trained on the COCO dataset~\cite{tsung-yi_lin_microsoft_2014} and fine-tuned on our cell detection datasets to leverage transfer learning, accelerate training and reduce the amount of data required for training.

Models were trained for 200 epochs using default Ultralytics augmentation settings, which include mosaic augmentation, mixup augmentation, and color space augmentations (HSV) to enhance model robustness and generalization.
An early stopping mechanism with a patience of 35 epochs was employed to prevent overfitting and optimize training time.
Training was performed on a single NVIDIA GeForce RTX 3090 GPU with 24GB of memory.

Model performance was evaluated on the held-out test set of 16,758 real images.
Performance metrics included mean Average Precision (mAP) at IoU thresholds of 0.5 (mAP@50), 0.75 (mAP@75), and averages across 0.5 to 0.95 (mAP@50:95).
mAP\@50 was chosen as the primary metric for performance comparison, as it provides a robust measure of detection accuracy while being less sensitive to precise boundary delineation, which is less critical in our single cell detection task focused on identifying and localizing cells rather than detailed morphological analysis.
However, mAP\@75 and mAP\@50:95 were also reported to provide a more comprehensive performance profile across different IoU thresholds.

\subsection{Expert Survey for Image Realism}
\label{subsec:expert-survey-for-image-realism}
To subjectively assess the perceptual realism of the generated synthetic images, we conducted a survey involving 11 microscopy experts and biologists from Synentec GmbH.
These experts possess extensive experience in microscopy image analysis and cell biology, making them ideal judges of image realism in this domain.

Participants were presented with 30 randomly ordered images, consisting of 20 synthetic images generated by our diffusion model and 10 real brightfield microscopy images from our dataset, ensuring a balanced representation of both image types without revealing the true ratio to participants.
For each image, participants were asked to clarify it as either ``real'' or ``synthetic'' and to rate their confidence in their classification on a scale from 1 (low confidence) to 5 (high confidence).
Furthermore, if participants classified an image as ``generated'', they were prompted to provide a brief textual explanation detailing the visual cues or artifacts that led to their decision.

We analyzed the classification accuracy of the experts, calculating overall accuracy as well as accuracy separately for real and synthetic images.
We also collected and analyzed the textual explanations provided by participants for images they classified as generated.
These textual explanations were analyzed qualitatively to identify recurring themes and visual features that experts associated with synthetic images.

This comprehensive methodology, combining quantitative evaluations of object detection performance with subjective assessments of image realism by expert human observers, provides a robust framework for investigating the impact of synthetic data augmentation in brightfield microscopy cell detection.

### 2025-02-26 sec/4_experimentsandresults.tex
extend introduction, related works and methodology chapters

        \multirow[t]{7}{*}{YOLOv8s}   & \textit{scc\_real} & 0.8947                    & \cellcolor{blue!10}0.8095  & \cellcolor{blue!10}0.6557     \\

@@

                                      & \textit{scc\_add\_30}   & 0.8952               & 0.8115                     & 0.6536                    \\

@@

        \multirow[t]{7}{*}{YOLOv8m}   & \textit{scc\_real} & \cellcolor{blue!10}0.9038  & \cellcolor{blue!10}0.8203 & \cellcolor{blue!10}0.6637     \\

@@

                                      & \textit{scc\_add\_30}   & 0.9034                    & 0.8224                     & 0.6606                    \\

@@

        \multirow[t]{7}{*}{YOLOv8x}   & \textit{scc\_real} & \cellcolor{blue!10}0.9038  & \cellcolor{blue!10}0.8120 & \cellcolor{blue!10}0.6639     \\

@@

                                      & \textit{scc\_add\_30}   & 0.8942                    & 0.8212                     & 0.6639                    \\

@@

        \multirow[t]{7}{*}{YOLOv9c}   & \textit{scc\_real} & \cellcolor{blue!10}0.9047  & \cellcolor{blue!10}0.8215 & \cellcolor{blue!10}0.6645     \\

@@

                                      & \textit{scc\_add\_30}   & 0.8952                    & 0.8217                     & 0.6605                    \\

@@

        \multirow[t]{7}{*}{YOLOv9e}   & \textit{scc\_real} & 0.9037                     & \cellcolor{blue!10}0.8201 & 0.6629     \\
                                      & \textit{scc\_10}   & \cellcolor{blue!10}0.9041  & 0.8141                    & \cellcolor{blue!10}0.6650 \\

@@

                                      & \textit{scc\_add\_30}   & 0.8952                    & 0.8209                     & 0.6638                    \\
                                      & \textit{scc\_add\_50}   & 0.9031                    & 0.8188                     & 0.6651                    \\

>>>>

        \multirow[t]{7}{*}{YOLOv8s}   & \textit{scc\_real}      & 0.8947                     & 0.8095                     & \cellcolor{blue!10}0.6557  \\

@@

                                      & \textit{scc\_add\_30}   & 0.8952                     & \cellcolor{blue!10}0.8115  & 0.6536                     \\

@@

        \multirow[t]{7}{*}{YOLOv8m}   & \textit{scc\_real}      & \cellcolor{blue!10}0.9038  & 0.8203                     & \cellcolor{blue!10}0.6637  \\

@@

                                      & \textit{scc\_add\_30}   & 0.9034                     & \cellcolor{blue!10}0.8224  & 0.6606                     \\

@@

        \multirow[t]{7}{*}{YOLOv8x}   & \textit{scc\_real}      & \cellcolor{blue!10}0.9038  & 0.8120                     & \cellcolor{blue!10}0.6639  \\

@@

                                      & \textit{scc\_add\_30}   & 0.8942                     & \cellcolor{blue!10}0.8212  & \cellcolor{blue!10}0.6639  \\

@@

        \multirow[t]{7}{*}{YOLOv9c}   & \textit{scc\_real}      & \cellcolor{blue!10}0.9047  & 0.8215                     & \cellcolor{blue!10}0.6645  \\

@@

                                      & \textit{scc\_add\_30}   & 0.8952                     & \cellcolor{blue!10}0.8217  & 0.6605                     \\

@@

        \multirow[t]{7}{*}{YOLOv9e}   & \textit{scc\_real}      & 0.9037                     & 0.8201                     & 0.6629                     \\
                                      & \textit{scc\_10}        & \cellcolor{blue!10}0.9041  & 0.8141                     & 0.6650                     \\

@@

                                      & \textit{scc\_add\_30}   & 0.8952                     & \cellcolor{blue!10}0.8209  & 0.6638                     \\
                                      & \textit{scc\_add\_50}   & 0.9031                     & 0.8188                     & \cellcolor{blue!10}0.6651  \\
